
# Reproducible Research: Peer assessment 1
### Ronald Miller 
### July 2015

==============================================================

## Introduction

This document presents the results of peer assessments 1 of course [Reproducible Research](https://class.coursera.org/repdata-004) on [coursera](https://www.coursera.org). This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day. 

## Assignment

This assignment will be described in multiple parts. You will need to write a report that answers the questions detailed below. Ultimately, you will need to complete the entire assignment in a single R markdown document that can be processed by knitr and be transformed into an HTML file. Throughout your report make sure you always include the code that you used to generate the output you present. When writing code chunks in the R markdown document, always use echo = TRUE so that someone else will be able to read the code. This assignment will be evaluated via peer assessment so it is essential that your peer evaluators be able to review the code for your analysis. For the plotting aspects of this assignment, feel free to use any plotting system in R (i.e., base, lattice, ggplot2) Fork/clone the GitHub repository created for this assignment. You will submit this assignment by pushing your completed files into your forked repository on GitHub. The assignment submission will consist of the URL to your GitHub repository and the SHA-1 commit ID for your repository state.

NOTE: The GitHub repository also contains the dataset for the assignment so you do not have to download the data separately.

## Prepare the R environment

First, we set echo equal a **TRUE** and results equal a **'hold'** as global options for this document.  
```{r set_options}
library(knitr)
opts_chunk$set(echo = TRUE, results = 'hold')
```

### Load required libraries
```{r call_libraries}
library(data.table)
library(ggplot2) 
```


## Loading and preprocessing the data

The following statement is used to load the data using `read.csv()`.

**Note**: It is assumed that the file activity.csv is in the current working directory. File can be downloaded from [here](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip)

```{r read_data}
rdata <- read.csv('activity.csv', header = TRUE, sep = ",", colClasses=c("numeric", "character", "numeric"))
```

### tidy the data or preprocess the data

We convert the **date** field to `Date` class and **interval** field to `Factor` class.

```{r tidy_data}
rdata$date <- as.Date(rdata$date, format = "%Y-%m-%d")
rdata$interval <- as.factor(rdata$interval)
```

Now, let us check the data using `str()` method:

```{r check_data}
str(rdata)
```

## What is mean total number of steps taken per day?

We proceed by calculating the total steps per day ignoring the missing values 

```{r pre_calc_stepsperday}
steps_per_day <- aggregate(steps ~ date, rdata, sum)
colnames(steps_per_day) <- c("date","steps")
head(steps_per_day)
```

1. Now we make a histogram of the total number of steps taken per day, plotted with appropriate bin interval.

```{r histo}
#png(filename = "Histogram1.png", width = 480, height = 480, units = "px")
ggplot(steps_per_day, aes(x = steps)) + geom_histogram(fill = "orange", binwidth = 1000) + 
        labs(title="Histogram of Steps Taken per Day", 
        x = "Number of Steps per Day", y = "Number of times in a day(Count)") + theme_bw() 
#dev.off()
```

2. Now we calculate the ***mean*** and ***median*** of the number of steps taken per day.

```{r meanmedian}
steps_mean   <- mean(steps_per_day$steps, na.rm=TRUE)
steps_median <- median(steps_per_day$steps, na.rm=TRUE)
```

The mean is **`r format(steps_mean,digits = 8)`** and median is **`r format(steps_median,digits = 8)`**.

## What is the average daily activity pattern?

We calculate the aggregation of steps by intervals of 5-minutes and convert the intervals as integers and save them in a data frame called `steps_per_interval`.

```{r steps_interval}
steps_per_interval <- aggregate(rdata$steps, by = list(interval = rdata$interval), FUN=mean, na.rm=TRUE)
steps_per_interval$interval <- as.integer(levels(steps_per_interval$interval)[steps_per_interval$interval])
        colnames(steps_per_interval) <- c("interval", "steps")
```


1. We make the plot with the time series of the average number of steps taken (averaged across all days) versus the 5-minute intervals:


```{r plot_time_series}
#png(filename = "TimeSeries.png", width = 480, height = 480, units = "px")
ggplot(steps_per_interval, aes(x=interval, y=steps)) + geom_line(color="blue", size=1) +  
        labs(title="Average Daily Activity Pattern", x="Interval", y="Number of steps") +  
        theme_bw()
#dev.off()
```


2. Now, we find the 5-minute interval with the containing the maximum number of steps:

```{r max_interval}
max_interval <- steps_per_interval[which.max(steps_per_interval$steps),]
```

The interval **`r max_interval$interval`** has a maximum of **`r round(max_interval$steps)`** steps.

## Imputing missing values:

### 1. Total number of missing values:

The total number of missing values is calculated in r using

```{r tot_na_value}

missing_vals <- sum(is.na(rdata$steps))

```

where the total number of ***missing values*** are **`r missing_vals`**.

### 2. Strategy for filling in all of the missing values in the dataset

To populate missing values, we choose to replace them with the mean value at the same interval across days. We create a function `na_fill(data, pervalue)` which the `data` arguement is the `rdata` data frame and `pervalue` arguement is the `steps_per_interval` data frame.

```{r fill_na}
na_fill <- function(data, pervalue) {
        na_index <- which(is.na(data$steps))
        na_replace <- unlist(lapply(na_index, FUN=function(idx){interval = data[idx,]$interval
                pervalue[pervalue$interval == interval,]$steps
        }))
        fill_steps <- data$steps
        fill_steps[na_index] <- na_replace
        fill_steps
}

rdata_fill <- data.frame(  
        steps = na_fill(rdata, steps_per_interval),  
        date = rdata$date,  
        interval = rdata$interval)
str(rdata_fill)
```

We check that are there any missing values remaining or not

```{r check_empty}
sum(is.na(rdata_fill$steps))
```

whereby a zero output is indicative of  ***NO MISSING VALUES***.

### 3. A histogram of the total number of steps taken each day

Now let us plot a histogram of the daily total number of steps taken, plotted with a bin interval of 1000 steps, after filling missing values.

```{r histo_fill}
fill_steps_per_day <- aggregate(steps ~ date, rdata_fill, sum)
colnames(fill_steps_per_day) <- c("date","steps")

##plotting the histogram
#png(filename = "Histogram2.png", width = 480, height = 480, units = "px")
ggplot(fill_steps_per_day, aes(x = steps)) + geom_histogram(fill = "blue", binwidth = 1000) + 
        labs(title="Histogram of Steps Taken per Day", 
        x = "Number of Steps per Day", y = "Number of times in a day(Count)") + theme_bw() 
#dev.off()

```

### Calculate and report the **mean** and **median** total number of steps taken per day.

```{r meanmedian_fill}
steps_mean_fill   <- mean(fill_steps_per_day$steps, na.rm=TRUE)
steps_median_fill <- median(fill_steps_per_day$steps, na.rm=TRUE)
```

The mean is **`r format(steps_mean_fill,digits = 8)`** and median is **`r format(steps_median_fill,digits = 8)`**.

### Do these values differ from the estimates from the first part of the assignment?

The values do differ slightly

- **Before filling the data**
    1. Mean  : **`r format(steps_mean,digits = 8)`**
    2. Median: **`r format(steps_median,digits = 8)`**  
- **After filling the data**
    1. Mean  : **`r format(steps_mean_fill,digits = 8)`**
    2. Median: **`r format(steps_median_fill,digits = 8)`**

However, we see that the values between the data mean and median are equal after filling in the missing values with interval means

### What is the impact of imputing missing data on the estimates of the total daily number of steps?

We observe that while the mean value remains unchanged, the median value has shifted and virtual matches to the mean. By imputing the missing values, the peak of the distribution has increased but it has not negatively affected our predictions.   


## Are there differences in activity patterns between weekdays and weekends?

In order to compare the time periods we will need to perform the following

1. Create a column that represents the days of the week  
2. Create a table with two parts, one representing weekends (Saturday and Sunday) and the other weekdays (Monday through Friday).  
3. Calculate the average steps per interval for each data set.  
4. Plot the two data sets for comparison.  

```{r weekdays}
weekdays_steps <- function(data) {
    weekdays_steps <- aggregate(data$steps, by=list(interval = data$interval), FUN=mean, na.rm=T)
    weekdays_steps$interval <- as.integer(levels(weekdays_steps$interval)[weekdays_steps$interval])
    colnames(weekdays_steps) <- c("interval", "steps")
    weekdays_steps
}

data_by_weekdays <- function(data) {
    data$weekday <- as.factor(weekdays(data$date))
    weekend_data <- subset(data, weekday %in% c("Saturday","Sunday"))
    weekday_data <- subset(data, !weekday %in% c("Saturday","Sunday"))
    weekend_steps <- weekdays_steps(weekend_data)
    weekday_steps <- weekdays_steps(weekday_data)
    weekend_steps$dayofweek <- rep("weekend", nrow(weekend_steps))
    weekday_steps$dayofweek <- rep("weekday", nrow(weekday_steps))
    data_by_weekdays <- rbind(weekend_steps, weekday_steps)
    data_by_weekdays$dayofweek <- as.factor(data_by_weekdays$dayofweek)
    data_by_weekdays
}

data_weekdays <- data_by_weekdays(rdata_fill)
```

Below you can see the panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends:
```{r plot_weekdays}
#png(filename = "Comparison.png", width = 480, height = 480, units = "px")
ggplot(data_weekdays, aes(x=interval, y=steps)) + geom_line(color="blue") + 
        facet_wrap(~ dayofweek, nrow=2, ncol=1) + labs(x="Interval", y="Number of steps") +
        theme_bw()
#dev.off()

```

Weekends activities has more peaks over a hundred than weekday. This is probably due to the fact that activities on weekends has more intesnity due to more free leisure time, as opposed to during the week where work related routines may hinder the activities.

